Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-09 ind_files/01_Introduction.tex
Fix in-text citation spacing and paragraph spacing

More than 150 years have passed since Charles Darwin first published \textit{Origin} (in 1859) and Ernst Haeckel first coined the term ‘ecology’ (in 1866). Today, both ecology and evolution are incredibly interdisciplinary fields, borrowing techniques and ideas from diverse areas such as computer science, statistics, economics, dynamical systems, physics, and information theory. With this development has come a cornucopia of models that try to understand biological phenomena in the language of these borrowed tools and techniques. Though many of these ideas are used to accurately describe and analyze specific systems, there is also value to formulating general models in abstract terms that only incorporate a small number of `fundamental' processes and try to capture the `essence' of a biological pattern \citep{frank_natural_2012, vellend_theory_2016,luque_mirror_2021}. Such general organizing models are vital to theory-building \citep{luque_mirror_2021} because they help clarify conceptual similarities and unifying factors between apparently disparate modelling frameworks, helping us seamlessly translate essential ideas from one theoretical language to another.

\section{Idealization and generality in ecology}\label{idealization}
The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology \citep{kingsland_modeling_1985} and evolution \citep{provine_origins_2001}. One of the first mathematical idealizations in ecology came in the form of the logistic equation formulated by Pearl and Verhulst, and shortly thereafter, equations for the populations of interacting predator and prey species, put forth independently by Lotka and Volterra. These models were immediately controversial, and for good reason: Many ecologists felt that they were overly idealistic and neglected many important truths about real biological populations \citep{kingsland_modeling_1985}. These models were nevertheless quite good at predicting the patterns of such populations, and have proved valuable to the field of population ecology. Today, these models are viewed as `classical' and are regularly used even by hard-line empiricists, not because we believe them to be true in all their gory biological details, but because we recognize that they can be \emph{useful} despite being blatantly false generalizations. This is a single instance of a more general philosophical idea concerning the goals and ideals of science. As suggested by the epigraph at the beginning of the chapter, philosophers of science \citep{potochnik_idealization_2018} have recently argued that science does not seek truth, but instead seeks understanding. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses \citep{zuk_models_2018}. Relatively simple models of complex eco-evolutionary processes are therefore desirable as a way to shine light on these phenomena. To reiterate once more, the goal of such simple models is not truth (whatever that is), predictive power (as with models in physics), or detailed description (as with detailed individual-based simulations, very flexible statistical models, or machine learning), but \emph{understanding}. Since the world is complicated and humans are limited, such understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. It is important to remember at the outset that the models I speak about in this thesis may seem to be too general and may seem to ignore many biological details. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.\\
Vellend has recently argued that conceptual synthesis in community ecology requires ``shifting the emphasis away from an organizational structure based on the useful lines of inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns" \citep{vellend_theory_2016}. Vellend's assertion is based on the fact that population genetics has managed to come up with reasonably comprehensive theory due to its focus on the abstract `high-level' processes of selection, mutation, drift, and gene flow instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on specific `low-level' processes such as predation rate, limiting resources ($R^*$), storage effects, priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table 5.1 in \citep{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that community ecology `is a mess'. Vellend proposes organizing ecological models according to the `high-level' processes of selection, ecological drift (demographic stochasticity), dispersal, and speciation.\\ Of course, no such general organization will be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, I believe the act of looking beyond the low-level processes present in biological systems and categorizing theories, models, and concepts in terms of a small number of (blurry-eyed) fundamental high-level processes provides a powerful unifying tool to organize concepts in biology.

@@

In population genetics, the relevant high-level forces are the standard evolutionary forces of natural selection, genetic drift, dispersal, and mutation. The description of evolution in terms of these forces was first laid out in formal mathematical terms during the Modern Synthesis by authors such as Wright, Fisher, and Haldane, an extremely successful venture that unified two major schools of thought --- Mendelian genetics and Darwinian evolution --- that were, at the time, thought to be incompatible \citep{provine_origins_2001}.\\
Classical population genetics regarded forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, the evolutionary play famously unfolds in the ecological theatre \citep{hutchinson_ecological_1965}. Thus, quantities like fitness are not truly fundamental but instead emerge as the net result of various ecological interactions, tradeoffs, and constraints \citep{metz_how_1992}, a fact that can have important consequences for evolution \citep{coulson_putting_2006, kokko_can_2017}. Trying to understand such `eco-evolutionary dynamics' has sprouted a rich body of theoretical literature under the broad heading of `evolutionary ecology' and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations \citep{brown_why_2016}. Eco-evolutionary population dynamics can (very) broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures (such as evolutionary game theory and classic population genetics) as special cases \citep{page_unifying_2002, queller_fundamental_2017, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time \citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology \citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner \citep{ lehtonen_price_2020, luque_mirror_2021}\\
One of the general guiding principles of much of this mathematization has been the assumption that incorporating the reality of finite population sizes into models leads to no major qualitative differences in behavior, only `adding noise' or `blurring out' the predictions of simpler infinite population models \citep{page_unifying_2002}. Consequently, several major theoretical frameworks in the field, such as adaptive dynamics, are explicitly formulated in deterministic terms at the infinite population size limit. However, this assumption is largely unjustified, and since populations in the real world are finite and stochastic, checking whether stochastic models differ from their deterministic analogs is vital to furthering our understanding of the fundamentals of population biology \citep{hastings_transients_2004, coulson_skeletons_2004, shoemaker_integrating_2020}.Today, we increasingly recognize that incorporating the finite and stochastic nature of the real world routinely has much stronger consequences than simply `adding noise' to deterministic expectations \citep{boettiger_noise_2018}, with important consequences for both ecological \citep{schreiber_does_2022} and evolutionary \citep{delong_stochasticity_2023} theory. In ecology and evolution, stochastic models often do not exhibit phenomena that occur in deterministic models \citep{proulx_what_2005, johansson_will_2006, claessen_delayed_2007,  wakano_evolutionary_2013, debarre_evolutionary_2016, johnson_two-dimensional_2021}, exhibit phenomena that do not occur in deterministic models \citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015, veller_drift-induced_2017, delong_stochasticity_2023}, and can even completely reverse the predictions of deterministic models \citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,constable_demographic_2016,mcleod_social_2019}. Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics \citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics \citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Importantly, the Price equation itself is usually formulated in a deterministic, dynamically-insufficient manner (but see \cite{rice_stochastic_2008} for a stochastic formulation of the Price equation). Since real-life populations are stochastic, finite, and of non-constant population size, it is thus imperative that we develop a theoretical framework that can handle such systems directly, instead of only working with deterministic, infinite-population approximations.\\
Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton' \citep{coulson_skeletons_2004} in an ad-hoc fashion, can lead to nonsensical predictions and inconsistencies \citep{strang_how_2019}. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are self-consistent \citep{strang_how_2019}, much more natural \citep{black_stochastic_2012}, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model \citep{strang_how_2019}. Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons \citep{metcalf_why_2007,geritz_mathematical_2012} because `all paths to fitness lead through demography' \citep{metcalf_all_2007}. In other words, since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes and provide a more fundamental mechanistic description of the relevant evolutionary forces and population dynamics \citep{doebeli_towards_2017}.

@@

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, use the system size expansion to arrive at a continuous description of change in trait frequencies as an SDE under mild assumptions on the functional forms of the birth and death rates. Unlike many classic stochastic formulations in evolutionary theory \citep{fisher_genetical_1930,wright_evolution_1931,moran_random_1958,crow_introduction_1970, lande_natural_1976}, I do not assume a fixed (effective) population size and instead allow the total population size to fluctuate over time. I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, the Price equation), thus establishing the microscopic basis of well-known equations from stochastic first principles. I also illustrate some general predictions that can be made using the weak noise approximation for the sake of completeness. While the mathematics of section \ref{sec_nD_processes} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see \cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach as applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapter \ref{sec_nD_processes} can be thought of as a tutorial and conceptual synthesis - For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions; For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution. Chapter \ref{chap_infD_processes} introduces a function-valued process\footnote{This is really a measure-valued process but we will ignore this in the interest of accessibility} to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research \citep{walsh_introduction_1986,carmona_stochastic_1999,da_prato_stochastic_2014,balan_gentle_2018} and the mathematics itself is thus still far from settled. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, has never been presented in full generality either mathematically or biologically. Mathematically, chapter \ref{chap_infD_processes} presents heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits \citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} by generalizing the idea of a system size expansion of density-dependent (finite-dimensional) birth-death processes to the infinite-dimensional case using the notion of functional differentiation. Biologically, Chapter \ref{chap_infD_processes} provides a series of functional partial differential equations (`stochastic field equations') to describe the dynamics of one-dimensional quantitative traits in populations. Some concrete models are presented as examples in Appendix \ref{App_examples} for clarity regarding the major ideas.\\
Part \ref{part_summary} summarizes the major results of our formalism and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of \cite{queller_fundamental_2017} or `unifying factors' in the sense of \cite{lion_theoretical_2018}. These equations reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same equations predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology and whose significance is only recently being recognized \citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}. Implications of noise-induced selection are also discussed in part \ref{part_summary}. Readers who are averse to or do not care for mathematical details can safely skip part \ref{part_theory} entirely and directly read part \ref{part_summary} for the major takeaways of this thesis, though I strongly recommend working through the formalism properly if possible.

@@

%Alternative reproductive tactics (ARTs) are discrete polymorphisms that occur in sympatry in one or both sexes of a species in response to intrasexual competition for mates. More precisely, two or more traits which are expressed in individuals of a given sex of a given species can be said to be ARTs if \citep{oliveira_alternative_2008}:

@@

%ARTs are widespread in the animal kingdom, and may be seen either in morphological traits, or as more complex behavioral polymorphisms (see \citep{oliveira_alternative_2008} for an overview). Most (if not all) species which exhibit ARTs only show a small number of polymorphisms (usually 2-3, 5 in a handful of cases, and very rarely 7-8) - \hl{I can maybe make a table for this: Somehow wasn't able to find one in the literature}. ARTs have obvious consequences for reproductive life history, and can strongly influence the ecology and evolutionary trajectory of organisms.

@@

%Trophic resource polymorphism is defined as ``the occurrence of discrete intraspecific morphs showing differential niche use, usually through differences in feeding biology and habitat use" \citep{skulason_resource_1995}. Species which exhibit trophic resource polymorphisms often show marked, discontinuous morphological variations adapted for specific niche usage. For example, arctic charr and african cichlid fishes both often exhibit discrete trophic polymorphisms in which different morphs (of the same species) are specialized for feeding at different strata of the lakes they live in \citep{recknagel_ecosystem_2017}. Such polymorphisms also generally occur in relatively small numbers in sympatry (see Table 1 in \citep{smith_evolutionary_1996}). They are relatively widespread in vertebrates, and are thought to have important evolutionary consequences by acting as the `substrate' for adaptive radiations \citep{smith_evolutionary_1996}. 

@@

%Isogamous organisms are sexual, but do not have gametes which can be classified into `male' and `female' based on size. Gametes of such species can, however, be divided into distinct `mating types' according to chemical self-incompatibility. There is substantial variation in the number of mating types present in a given species: While most organisms have only two mating types, this is by no means the rule \citep{phadke_rapid_2009,constable_rate_2018}, with a handful of organisms having 7-10 mating types, and one species of fungus (\textit{Schizophyllum commune}) exhibiting over 23,000 mating types. The study of mating types in isogamous species is ecologically and evolutionarily relevant, not only due to the intrinsic importance of mating and self-incompatibility, but due to the fact that isogamy is thought to be the evolutionary precursor to anisogamy: Thus, studying the factors governing the number of mating types in isogamous species may shed light on why anisogamous species have only two distinct types of gametes.

>>>>

More than 150 years have passed since Charles Darwin first published \textit{Origin} (in 1859) and Ernst Haeckel first coined the term ‘ecology’ (in 1866). Today, both ecology and evolution are incredibly interdisciplinary fields, borrowing techniques and ideas from diverse areas such as computer science, statistics, economics, dynamical systems, physics, and information theory. With this development has come a cornucopia of models that try to understand biological phenomena in the language of these borrowed tools and techniques. Though many of these ideas are used to accurately describe and analyze specific systems, there is also value to formulating general models in abstract terms that only incorporate a small number of `fundamental' processes and try to capture the `essence' of a biological pattern~\citep{frank_natural_2012, vellend_theory_2016,luque_mirror_2021}. Such general organizing models are vital to theory-building~\citep{luque_mirror_2021} because they help clarify conceptual similarities and unifying factors between apparently disparate modelling frameworks, helping us seamlessly translate essential ideas from one theoretical language to another.

\section{Idealization and generality in ecology}\label{idealization}
The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology~\citep{kingsland_modeling_1985} and evolution~\citep{provine_origins_2001}. One of the first mathematical idealizations in ecology came in the form of the logistic equation formulated by Pearl and Verhulst, and shortly thereafter, equations for the populations of interacting predator and prey species, put forth independently by Lotka and Volterra. These models were immediately controversial, and for good reason: Many ecologists felt that they were overly idealistic and neglected many important truths about real biological populations~\citep{kingsland_modeling_1985}. These models were nevertheless quite good at predicting the patterns of such populations, and have proved valuable to the field of population ecology. Today, these models are viewed as `classical' and are regularly used even by hard-line empiricists, not because we believe them to be true in all their gory biological details, but because we recognize that they can be \emph{useful} despite being blatantly false generalizations. This is a single instance of a more general philosophical idea concerning the goals and ideals of science. Som philosophers of science have recently argued that ``the epistemic goal of science is not truth, but understanding''~\citep{potochnik_idealization_2018}. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses~\citep{zuk_models_2018}. Relatively simple models of complex eco-evolutionary processes are therefore desirable as a way to shine light on these phenomena. To reiterate once more, the goal of such simple models is not truth (whatever that is), predictive power (as with models in physics), or detailed description (as with detailed individual-based simulations, very flexible statistical models, or machine learning), but \emph{understanding}. Since the world is complicated and humans are limited, such understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. It is important to remember at the outset that the models I speak about in this thesis may seem to be too general and may seem to ignore many biological details. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.

Vellend has recently argued that conceptual synthesis in community ecology requires ``shifting the emphasis away from an organizational structure based on the useful lines of inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns"~\citep{vellend_theory_2016}. Vellend's assertion is based on the fact that population genetics has managed to come up with reasonably comprehensive theory due to its focus on the abstract `high-level' processes of selection, mutation, drift, and gene flow instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on specific `low-level' processes such as predation rate, limiting resources ($R^*$), storage effects, priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table 5.1 in~\citep{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that community ecology `is a mess'. Vellend proposes organizing ecological models according to the `high-level' processes of selection, ecological drift (demographic stochasticity), dispersal, and speciation.

Of course, no such general organization will be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, I believe the act of looking beyond the low-level processes present in biological systems and categorizing theories, models, and concepts in terms of a small number of (blurry-eyed) fundamental high-level processes provides a powerful unifying tool to organize concepts in biology.

@@

In population genetics, the relevant high-level forces are the standard evolutionary forces of natural selection, genetic drift, dispersal, and mutation. The description of evolution in terms of these forces was first laid out in formal mathematical terms during the Modern Synthesis by authors such as Wright, Fisher, and Haldane, an extremely successful venture that unified two major schools of thought --- Mendelian genetics and Darwinian evolution --- that were, at the time, thought to be incompatible~\citep{provine_origins_2001}.

Classical population genetics regarded forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, the evolutionary play famously unfolds in the ecological theatre~\citep{hutchinson_ecological_1965}. Thus, quantities like fitness are not truly fundamental but instead emerge as the net result of various ecological interactions, tradeoffs, and constraints~\citep{metz_how_1992}, a fact that can have important consequences for evolution~\citep{coulson_putting_2006, kokko_can_2017}. Trying to understand such `eco-evolutionary dynamics' has sprouted a rich body of theoretical literature under the broad heading of `evolutionary ecology' and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations~\citep{brown_why_2016}. Eco-evolutionary population dynamics can (very) broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures (such as evolutionary game theory and classic population genetics) as special cases~\citep{page_unifying_2002, queller_fundamental_2017, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time~\citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology~\citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner~\citep{ lehtonen_price_2020, luque_mirror_2021}

One of the general guiding principles of much of this mathematization has been the assumption that incorporating the reality of finite population sizes into models leads to no major qualitative differences in behavior, only `adding noise' or `blurring out' the predictions of simpler infinite population models~\citep{page_unifying_2002}. Consequently, several major theoretical frameworks in the field, such as adaptive dynamics, are explicitly formulated in deterministic terms at the infinite population size limit. However, this assumption is largely unjustified, and since populations in the real world are finite and stochastic, checking whether stochastic models differ from their deterministic analogs is vital to furthering our understanding of the fundamentals of population biology~\citep{hastings_transients_2004, coulson_skeletons_2004, shoemaker_integrating_2020}.Today, we increasingly recognize that incorporating the finite and stochastic nature of the real world routinely has much stronger consequences than simply `adding noise' to deterministic expectations~\citep{boettiger_noise_2018}, with important consequences for both ecological~\citep{schreiber_does_2022} and evolutionary~\citep{delong_stochasticity_2023} theory. In ecology and evolution, stochastic models often do not exhibit phenomena that occur in deterministic models~\citep{proulx_what_2005, johansson_will_2006, claessen_delayed_2007,  wakano_evolutionary_2013, debarre_evolutionary_2016, johnson_two-dimensional_2021}, exhibit phenomena that do not occur in deterministic models~\citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015, veller_drift-induced_2017, delong_stochasticity_2023}, and can even completely reverse the predictions of deterministic models~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,constable_demographic_2016,mcleod_social_2019}. Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Importantly, the Price equation itself is usually formulated in a deterministic, dynamically-insufficient manner (but see~\cite{rice_stochastic_2008} for a stochastic formulation of the Price equation). Since real-life populations are stochastic, finite, and of non-constant population size, it is thus imperative that we develop a theoretical framework that can handle such systems directly, instead of only working with deterministic, infinite-population approximations.

Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton'~\citep{coulson_skeletons_2004} in an ad-hoc fashion, can lead to nonsensical predictions and inconsistencies~\citep{strang_how_2019}. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are self-consistent~\citep{strang_how_2019}, much more natural~\citep{black_stochastic_2012}, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model~\citep{strang_how_2019}. Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons~\citep{metcalf_why_2007,geritz_mathematical_2012} because `all paths to fitness lead through demography'~\citep{metcalf_all_2007}. In other words, since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes and provide a more fundamental mechanistic description of the relevant evolutionary forces and population dynamics~\citep{doebeli_towards_2017}.

@@

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner.

Chapter \ref{chap_BD} deals with the evolution of discrete traits. In this case, the system is finite-dimensional, since we can completely specify the state of the system by simply listing out the number of individuals of each type in a vector. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, use the system size expansion to arrive at a continuous description of change in trait frequencies as an SDE under mild assumptions on the functional forms of the birth and death rates. Unlike many classic stochastic formulations in evolutionary theory~\citep{fisher_genetical_1930,wright_evolution_1931,moran_random_1958,crow_introduction_1970, lande_natural_1976}, I do not assume a fixed (effective) population size and instead allow the total population size to fluctuate over time. I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, the Price equation), thus establishing the microscopic basis of well-known equations from stochastic first principles. I also illustrate some general predictions that can be made using the weak noise approximation for the sake of completeness. While the mathematics of section \ref{sec_nD_processes} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach as applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapter \ref{chap_BD} also serves as a tutorial and technical introduction to the major ideas - For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions; For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution.

Chapter \ref{chap_infD_processes} introduces a function-valued process\footnote{This is really a measure-valued process but we will pretend we don't know this in the interest of accessibility} to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{walsh_introduction_1986,carmona_stochastic_1999,da_prato_stochastic_2014,balan_gentle_2018} and the mathematics itself is thus still far from settled. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues~\citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, has never been presented in full generality either mathematically or biologically. Mathematically, chapter \ref{chap_infD_processes} presents heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits~\citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} by generalizing the idea of a system size expansion of density-dependent (finite-dimensional) birth-death processes to the infinite-dimensional case using the notion of functional differentiation. Biologically, Chapter \ref{chap_infD_processes} provides a series of functional partial differential equations (`stochastic field equations') to describe the dynamics of one-dimensional quantitative traits in populations. Some concrete models are presented as examples in Appendix \ref{App_examples} for clarity regarding the major ideas.

Part \ref{part_summary} summarizes the major results of our formalism and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of~\cite{queller_fundamental_2017} or `unifying factors' in the sense of~\cite{lion_theoretical_2018}. These equations reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same equations predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology and whose significance is only recently being recognized~\citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}. Implications of noise-induced selection are also discussed in part \ref{part_summary}. Readers who are averse to or do not care for mathematical details can safely skip part \ref{part_theory} entirely and directly read part \ref{part_summary} for the major takeaways of this thesis, though I strongly recommend working through the formalism properly if possible.

@@

%Alternative reproductive tactics (ARTs) are discrete polymorphisms that occur in sympatry in one or both sexes of a species in response to intrasexual competition for mates. More precisely, two or more traits which are expressed in individuals of a given sex of a given species can be said to be ARTs if~\citep{oliveira_alternative_2008}:

@@

%ARTs are widespread in the animal kingdom, and may be seen either in morphological traits, or as more complex behavioral polymorphisms (see~\citep{oliveira_alternative_2008} for an overview). Most (if not all) species which exhibit ARTs only show a small number of polymorphisms (usually 2-3, 5 in a handful of cases, and very rarely 7-8) - \hl{I can maybe make a table for this: Somehow wasn't able to find one in the literature}. ARTs have obvious consequences for reproductive life history, and can strongly influence the ecology and evolutionary trajectory of organisms.

@@

%Trophic resource polymorphism is defined as ``the occurrence of discrete intraspecific morphs showing differential niche use, usually through differences in feeding biology and habitat use"~\citep{skulason_resource_1995}. Species which exhibit trophic resource polymorphisms often show marked, discontinuous morphological variations adapted for specific niche usage. For example, arctic charr and african cichlid fishes both often exhibit discrete trophic polymorphisms in which different morphs (of the same species) are specialized for feeding at different strata of the lakes they live in~\citep{recknagel_ecosystem_2017}. Such polymorphisms also generally occur in relatively small numbers in sympatry (see Table 1 in~\citep{smith_evolutionary_1996}). They are relatively widespread in vertebrates, and are thought to have important evolutionary consequences by acting as the `substrate' for adaptive radiations~\citep{smith_evolutionary_1996}. 

@@

%Isogamous organisms are sexual, but do not have gametes which can be classified into `male' and `female' based on size. Gametes of such species can, however, be divided into distinct `mating types' according to chemical self-incompatibility. There is substantial variation in the number of mating types present in a given species: While most organisms have only two mating types, this is by no means the rule~\citep{phadke_rapid_2009,constable_rate_2018}, with a handful of organisms having 7-10 mating types, and one species of fungus (\textit{Schizophyllum commune}) exhibiting over 23,000 mating types. The study of mating types in isogamous species is ecologically and evolutionarily relevant, not only due to the intrinsic importance of mating and self-incompatibility, but due to the fact that isogamy is thought to be the evolutionary precursor to anisogamy: Thus, studying the factors governing the number of mating types in isogamous species may shed light on why anisogamous species have only two distinct types of gametes.
